\cleardoublepage
\chapter{Discussion, Future Work and Conclusions}
\label{dc} Wrap up (in separate sections, or integrated in one final
section):

\begin{description}
  \item[Discussion] Explain in the discussion section of your document information
presented in the results section, commenting on significant data and
experience produced by the study.
  \item[Conclusions] Include a conclusion as the final part of the body of your document.
    Because some readers of documents, particularly managers,
    will sometimes not read the entire document but, instead, focus on the conclusion,
    this part of the document should summarize all essential information necessary for your audience's purpose.
    In your conclusion:
    \begin{itemize}
      \item Relate your findings to the general problem and any specific objectives posed in your introduction.
      \item Summarize clearly what the report does and does not demonstrate.

      \item Include specific recommendations for action or for further research.
      Sometimes these recommendations will constitute a separate section of a document.

    \end{itemize}
  \item[Recommendations] Include appropriate and specific recommendations as part of your conclusion or,
  in feasibility and recommendation reports, as a separate section preceding the conclusion.

Many types of scientific and technical documents conclude by
pointing to further action. Research reports often recommend further
studies to confirm tentative explanations or to answer questions
presented in the discussion section. Feasibility and recommendation
reports always have one or more specific recommendations as the
principal aim of the document.

Recommendations should always be specific and appropriate to the
document's audience. Separate each specific recommendation. Often
authors present recommendations in bulleted or numbered lists.
Organize recommendations either in the order of importance or in the
logical order of development.
\end{description}
%%% Local Variables:
%%% TeX-master: "main"
%%% End:
