\cleardoublepage
\chapter{Introduction}
\label{intro} Computer games are a major part of the entertainment industry
these days, with major titles having budgets matching Hollywood movies. Even so,
there is a lot of potential for improvement on their appeal. Artificial
Intelligence is an area where games can improve substantially as limitations of
the past are disappearing, creating games that seem more realistic and
interesting. Increases in the computing power of the average computer has made
it possible to take into use techniques previously thought too computationally
expensive. Machine Learning has the power to create NPC characters that seem
intelligent and can add more realism to games, and as such make the game more
satisfying to the player.

\section{Motivation}
\label{sec:motivation}

Video games are a massive industry with budgets comparing to the ones of
Hollywood movies, with Rockstar's GTA 4 having an estimated production budget of
\$100 million. It follows that video games are a big part of people's regular
entertainment, with young people spending on average two hours per day on video
games. \citep{norway08:video_games}

Since the inception of the game industry, game producers have been heavily
focused on creating more and more audio-visually impressive games, and this
development has reached a point where the improvements made can only be
minor. During this process, the development of the AI for game characters has
been only as necessary, to meet the demand of players for characters that are as
sophisticated as the graphics. \citep{champandard2003ai}

Until the late 90s there had not been much focus on improving the AI utilized by
the characters in games\cite{tozour2002evolution}, but this is shifting rapidly,
as games focus more on intelligent characters, using it as their latest and
greatest selling point. There have even been entire game series that focus
entirely on AI, such as \emph{Creatures} \citep{grand1997creatures}, \emph{Black
  \& White} \citep{lionhead01:black_white} and \emph{The Sims}
\citep{maxis00:sims}.

Despite this, the majority of game AI still uses simple techniques that have
been the cornerstone of game AI since its early days. Simple scripted rules and
Finite State Machines still dominate the implementation methods due to their
simplicity, performance and ability to function ``well enough'' to implement
simple AIs. \citep{woodcock2001game} This creates AI that, while entertaining
and believable in most situations, is ultimately limited by the forethought of
the AI programmers, and is easily outmatched by most experienced players.

In recent times Massively Multiplayer Online Games have emerged and become
immensely popular, through titles such as \emph{World of Warcraft}, and games
have grown to have a high focus on Player versus Player multiplayer, with
popular shooter game series having either no single player campaign, or an
extremely short one. This trend gives a strong indication that players are
looking for more intelligent opponents and more varied game play, and that this
cannot be provided through the single player opponents in the games.

To alleviate the limitations of simple, static AI opponents, machine learning
could be utilized. ML could create more engaging and interesting computer
opponents, and enables the system to scale up to more complex worlds and
situations. \citep{funge2004artificial} In the same vein, it is believed that
utilizing evolutionary computing can enable the creating of computer opponents
that are interesting and interactive, and as such
entertaining. \citep{yannakakis2005ai}

This thesis will explore if ADATE \citep{olsson1994inductive}, a system for
inductive programming (an offshoot of evolutionary computing), can be
employed to generate entertaining AI opponents. Additional challenges consist of
finding a way to quantify the entertainment value of the generated agents, and
the creation of a game framework to support the research. Towards this end,
there are a set of objectives and questions (shown in the following section)
that need to be addressed to reach our goal.

\section{Objectives and Research Questions}
\label{sec:objective}

Given the motivation for this thesis, there are a selection of questions to be
addressed to further our objectives. The questions that will be answered in this
thesis are as follows.

\begin{enumerate}
\item How can a player's satisfaction in a game (entertainment) be measured
  quantitatively?
\item Can the use of evolutionary computing increase a game's interest and
  ultimately create a more interesting game than static AI?
\end{enumerate}

\begin{enumerate}
\item Establish a metric for measuring a game opponent's interest level to the
  player.
\item Adapt this metric to the ADATE platform and decide upon a test bed suited
  to the platform and metric. \marginpar{maybe change this when a test bed is located?}
\item Use the platform to enhance the entertainment level of the computer
  opponents in the test bed.
\end{enumerate}

\section{Summary}
\label{sec:summary}


\section{Thesis layout}
\label{sec:thesis-layout}




%%% Local Variables:
%%% TeX-master: "main"
%%% End:

% LocalWords:  Warcraft
