\cleardoublepage
\chapter{Introduction}
\label{intro} The introduction to your document should lead your readers into
your paper and give them an idea of what to expect. It should not be simply a
restatement of the abstract even though it will contain some of the same
material.

Introductions often do the following:
\begin{itemize}
  \item State the subject of your document as clearly as possible

  \item Define the problem you are addressing, your approach to the problem, and
    why this problem is important
  \item State the purpose of your document
  \item Define the scope of your document
  \item Provide necessary and relevant background information
  \item Give an outline of the rest of the document
\end{itemize}


Because the introduction leads your reader into your document, try to begin with
a general statement about the topic before moving on to specific issues. This
strategy will help make the topic accessible to your readers, especially those
who are not specialists in the field. Illustrations, like Figure~, often help to
introduce the reader to the problem.

More stuff in Chapter~\ref{rel}.

---

Computer games are a major part of the entertainment industry these days, with
major titles having budgets matching Hollywood movies. Even so, there is a lot
of potential for improvement on their appeal. Artificial Intelligence is an area
where games can improve substantially as limitations of the past are
disappearing, creating games that seem more realistic and interesting. Increases
in the computing power of the average computer has made it possible to take into
use techniques previously thought too computationally expensive. Machine
Learning has the power to create NPC characters that seem intelligent and can
add more realism to games, and as such make the game more satisfying to the
player.

\section{Motivation}
\label{sec:motivation}

Computer games are a massive industry \marginpar{Find revenue numbers} and a big
part of many people's regular entertainment \marginpar{how many people play
  games?}.

Since its inception, game producers have been heavily focused on creating more
and more graphically impressive games, and has reached a point where the
improvements made can only be minor.

There has not been much focus on improving the AI utilized by the characters in
games until recently\cite{tozour2002evolution}.

Machine learning could be utilized to create more engaging and interesting
computer opponents.

This thesis will explore if ADATE, a system for doing machine learning, can be
employed to generate entertaining AI opponents.

\section{Objective}
\label{sec:objective}

\section{Summary}
\label{sec:summary}

\section{Thesis layout}
\label{sec:thesis-layout}




%%% Local Variables:
%%% TeX-master: "main"
%%% End:
